Decision Assistance and Steering in Health Insurance

### 2021-06-14 finals/paper.tex
update draft

\vspace{-2ex}
\begin{abstract}
\noindent In many markets, consumers have access to an agent or some other intermediary to aid in decision making. The influence of such agents on consumer decisions, and the firm's influence on the behaviors of its agents, have important implications for the efficiency of these markets. In this paper, we study the welfare effects of decision assistance and firm steering in health insurance using the population of enrollments from the California Affordable Care Act (ACA) exchange from 2014 to 2019. We find that enrollees are nearly 2x more likely to select a silver plan when using some form of decision assistance and 30\% less likely to select a ``dominated'' plan. Preliminary results suggest that firms have considerable ability to steer consumers' decisions using agent commissions, where a \$1 increase in commissions paid to agents has the same effect as a \$2 or \$3 increase in plan premiums.
\end{abstract}

@@

Our empirical analysis proceeds as follows. First, we examine whether decision assistance affects plan choice. Our estimand of interest is the average treatment effect on the treated (ATT) for enrollees receiving decision assistance. We employ as our estimator an augmented inverse propensity weighted nested logit regression, as discussed in more detail in Section \ref{sec:causal}, and we limit the analysis to newly enrolled households. We estimate that such households are 50\% more likely to select a silver plan and 20\% less likely to remain uninsured when using some form of decision assistance. We also find strong evidence of heterogeneous effects depending on the type of assistance used, wherein households using public decision assistance ...

@@

Third,

We contribute to two distinct literatures. First, our examination of decision assistance and health insurance relates to the literature on the role of experts in decision making.

A related literature considers the effects of recommendations, sometimes automated via an algorithmic recommendation, on decisions. \cite{bundorf2019}, for example, show that offering additional information alongside an algorithmic expert recommendation significantly affects an individual's choice of prescription drug plans.

Second, we contribute to the literature examining the prevalence and underlying mechanisms of poor decision making in health insurance. In the employer-provided insurance market, \cite{liu2021} documents that over half of employers offer a dominated plan among their menu of insurance options available to their employees, and \cite{sinaiko2011} and \cite{bhargava2017} find that a large share of employees regularly select such dominated insurance plans. \cite{abaluck2011}, \cite{ketcham2012}, \cite{zhou2012}, \cite{heiss2013}, \cite{gruber2017}, \cite{ho2017rand} and many others similarly examine poor decisions when selecting a Medicare prescription drug (Part D) plan. The consensus from this literature is that individuals regularly make \textit{ex ante} very bad health insurance decisions, often selecting plans that are either financially dominated by other plan options or most likely suboptimal given the enrollee's existing medical conditions.

Specifically in the context of the ACA, \cite{orzol2018} find that an ``Enroll America,'' a national non-profit organization designed to expand health insurance enrollment in the U.S., significantly increased enrollment in the ACA exchange plans. \cite{myerson2019} also finds that federally funded assistance programs significantly increased the probability of enrollment in Medicaid, although they found little effect of such federal funding on enrollment in exchange plans. \cite{sommers2015} similarly show that access to navigators is highly correlated with Medicaid enrollment, mirroring earlier findings from \cite{aizer2003} in her study of Medicaid enrollment and application assistants in California.

\cite{aizawa2021}

While not definitive, a common theme in this literature is that poor health insurance decisions are due in-part to an inability or unwillingness of enrollees to accurately evaluate their plan options \citep{bhargava2017, hanoch2009}. Even still, \cite{ketcham2012} find that any such poor decision making will tend to improve over time, thereby partially alleviating the need for any explicit intervention.

@@

We see from Table~\ref{tab:summary-stats} that enrollees using decision assistance are of lower income and larger household size. Among types of plans purchased, we see little difference between those with and without decision assistance, with the exception of Kaiser plans for which individuals using decision assistance are less likely to purchase. Overall, the silver tier is the most commonly selected option, and this is particularly true for those using decision assistance. Enrollees with decision assistance are also much more likely to select an HMO and relatively more likely to select a PPO.

@@

@@

\begin{table}[h!]
\centering
\footnotesize
\begin{minipage}[h]{6in}
\caption[caption]{\textbf{Summary Statistics for Covered California Enrollees}\footnote{Statistics based on the population of households enrolled in a Covered California plan at any point from 2014-2019.}}
\centerline{%
    \input{tables/summary_stats.tex}
}
\label{tab:summary-stats}
\end{minipage}
\end{table}

>>>>

\vspace{-2ex}
\begin{abstract}
\noindent In many markets, consumers have access to an agent or some other intermediary to aid in decision making. The influence of such agents on consumer decisions, and the firm's influence on the behaviors of its agents, have important implications for the efficiency of these markets. In this paper, we study the welfare effects of decision assistance and firm steering in health insurance using the population of enrollments from the California Affordable Care Act (ACA) exchange from 2014 to 2019. We find that enrollees 30\% more likely to select a silver plan when using some form of decision assistance and 30\% less likely to select a ``dominated'' plan. Preliminary results suggest that firms have considerable ability to steer consumers' decisions using agent commissions, where a \$1 increase in commissions paid to agents has the same effect as a \$2 or \$3 increase in plan premiums.
\end{abstract}

@@

Our empirical analysis proceeds as follows. First, we examine whether decision assistance affects plan choice. Our estimand of interest is the average treatment effect on the treated (ATT) for enrollees receiving decision assistance. We employ as our estimator an augmented inverse propensity weighted nested logit regression, as discussed in more detail in Section \ref{sec:causal}, and we limit the analysis to newly enrolled households. We estimate that such households are just over 30\% more likely to select a silver plan, the majority of which comes from a reduction in the probability of selecting a bronze plan.

@@

Third, we incorporate insurer commission payments alongside premium-setting and demand estimation as part of a structural model of insurer steering. Results from this analysis are pending; however, preliminary analysis of commission payments and plan choice suggests that commissions are a larger factor in plan choice than premiums. For example, we estimate that a \$1 increase in commissions paid to private insurance agents is equivalent to a \$2 to \$3 increase in premiums. In other words, private insurance agents appear to have significant influence on plan choice, with commission payments having 2x the effect of premiums.

We contribute to two distinct literatures. First, our examination of decision assistance and health insurance relates to the literature on the role of experts in decision making.\footnote{A broader literature considers ``persuasion'', which includes the role of advertising and media reports. We are interested in the more narrow literature specifically dealing with decision assistance in the form of third-party ``experts'' or advice. See \cite{dellavigna2010} for a general survey of the empirical evidence on persuasion.} Theoretical models of the role of experts and advice in a principal-agent framework include \cite{szalay2005}, \cite{alonzo2008}, and \cite{armstrong2010}. In an experimental setting, \cite{schotter2003} finds that individuals are much more likely to follow the advice of an expert and that such behavior is welfare-improving. In education, \cite{borghans2015} find large effects of high school counselors on students' college field choices, and \cite{bettinger2012} estimate large effects of personal assistance on the probability of applying for and receiving college financial aid, with subsequently large effects on college attendance. \cite{finkelstein2019} similarly find large effects of both improved information and personal assistance on probability of enrollment in the Supplemental Nutrition Assistance Program among eligible households.

A related literature considers the effects of recommendations, sometimes automated via an algorithmic recommendation, on decisions. \cite{bundorf2019}, for example, show that offering additional information alongside an algorithmic expert recommendation significantly affects an individual's choice of prescription drug plans. \cite{kling2012} also estimates large effects of personalized cost information on plan switching in the Medicare Part D market. To our knowledge, the literature on experts and decision-making in the context of health insurance focuses on unbiased and often anonymous ``experts'' (e.g., an algorithm suggesting the best plan for an individual). Our setting is unique in that individuals have access to a public ``navigator,'' who receives no compensation for any plan choice, and a private insurance agent who receives a commission that varies by plan.

Second, we contribute to the literature examining the prevalence and underlying mechanisms of poor decision making in health insurance. In the employer-provided insurance market, \cite{liu2021} documents that over half of employers offer a dominated plan among their menu of insurance options available to their employees, and \cite{sinaiko2011} and \cite{bhargava2017} find that a large share of employees regularly select such dominated insurance plans. \cite{abaluck2011}, \cite{ketcham2012}, \cite{zhou2012}, \cite{heiss2013}, \cite{gruber2017}, \cite{ho2017rand} and many others similarly examine poor decisions when selecting a Medicare prescription drug (Part D) plan. The consensus from this literature is that individuals regularly make \textit{ex ante} very bad health insurance decisions, often selecting plans that are either financially dominated by other plan options or most likely suboptimal given the enrollee's existing medical conditions. While not definitive, a common theme in this literature is that poor health insurance decisions are due in-part to an inability or unwillingness of enrollees to accurately evaluate their plan options \citep{bhargava2017, hanoch2009}.

There is less consensus, however, in terms of how best to improve health insurance plan choice. For example, \cite{ketcham2012} find that observed poor health insurance decision making will tend to improve over time, thereby partially alleviating the need for any explicit intervention. In the context of the ACA, \cite{orzol2018} find that an ``Enroll America,'' a national non-profit organization designed to expand health insurance enrollment in the U.S., significantly increased enrollment in the ACA exchange plans. \cite{myerson2019} also finds that federally funded assistance programs significantly increased the probability of enrollment in Medicaid, although they found little effect of such federal funding on enrollment in exchange plans. \cite{sommers2015} similarly show that access to navigators is highly correlated with Medicaid enrollment, mirroring earlier findings from \cite{aizer2003} in her study of Medicaid enrollment and application assistants in California. \cite{aizawa2021} distinguish between ``public'' provision (via government advertising) versus private provision (via insurer advertising) on plan choice, where they find that both forms of advertising affect participation in the market and that private advertising has a particularly large effect on the intensive margin of plan choice; however, the welfare effects of private advertising are unclear due to business stealing and costs of advertising.

Our contribution to this literature is most closely related to \cite{aizawa2021} in that we consider both private and public decision assistance; however, we examine this in the context of one-on-one assistance rather than general exposure to advertising. We also consider the potential for private agents to improve health insurance decisions while also allowing for potential bias introduced through commissions (i.e., steering), and we compare outcomes from the use of private agents to those of public navigators. In this sense, our analysis follows in the spirit of \cite{ketcham2012} in that we envision private agents as helping to resolve an information problem, the costs of which include potential steering into sub-optimal plans and spillover effects of commission payments onto premiums. Whether the costs of such private agents exceed the benefits or are best borne by public agents without steering incentives is an empirical question which we pursue in Section \ref{sec:steering}.

@@

We see from Table~\ref{tab:summary-stats} that enrollees using decision assistance are of lower income and larger household size. Among types of plans purchased, we see relatively minor differences between those with and without decision assistance, with the exception of Kaiser plans for which individuals using decision assistance are less likely to purchase. Overall, the silver tier is the most commonly selected option, and this is particularly true for those using decision assistance. Enrollees with decision assistance are also slightly more likely to select a PPO relative to those without assistance.

@@

With regard to metal tier, Figure~\ref{fig:choice-metal} shows that individuals with decision assistance are about 16 percentage points more likely to select a silver tier plan. With 50\% of unassisted households selecting a silver plan, this reflects a relative increase of 32\% in the probability of selecting a silver plan. Most of this increase comes from the 50\% reduction in the probability of selecting a bronze plan, with a smaller (but still significant) reduction in selecting gold plans.

@@

\begin{table}[h!]
\centering
\footnotesize
\begin{minipage}[h]{6in}
\caption[caption]{\textbf{Summary Statistics for Covered California Enrollees}\footnote{Statistics based on the population of households enrolled in a Covered California plan at any point from 2014-2019. Household demographics and income summary statistics are based on all households, while summaries for insurer, metal tier, and network type are conditional on being insured.}}
\centerline{%
    \input{tables/summary_stats.tex}
}
\label{tab:summary-stats}
\end{minipage}
\end{table}
